A-maze of Natural Stories: Texts are comprehensible using the Maze task

### 2022-08-22 Papers/Paper/manuscript.Rmd
reorder models in figs, credit roles

---
title             : "A-maze of Natural Stories: Comprehension and surprisal in the Maze task"
shorttitle        : "A-maze of Natural Stories"
shortauthors: Veronica Boyce & Roger Levy
author: 
  - name          : "Veronica Boyce"
    affiliation   : "Stanford University"
    corresponding : yes    # Define only one corresponding author
    email         : "vboyce@stanford.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Investigation
      - Software
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Roger Levy"
    affiliation   : "Massachusetts Institute of Technology"
    role:
      - Conceptualization
      - Formal Analysis
      - Writing - Review & Editing

@@

classoption: [times, biblatex]
#
# Possible classoptions:
# - [times] for Times font (default if no option is chosen)
# - [cm] for Computer Modern font
# - [lucida] for Lucida font (not freely available)
# - [brill] open type font, freely downloadable for non-commercial use from
#   http://www.brill.com/about/brill-fonts; requires xetex
# - [charis] for CharisSIL font, freely downloadable from http://software.sil.org/charis/
# - for the Brill an CharisSIL fonts, you have to use the XeLatex
#   typesetting engine (not pdfLatex)
# - for headings, tables, captions, etc., Fira Sans is
#   used: https://www.fontsquirrel.com/fonts/fira-sans
# - [biblatex] for using biblatex (the default is natbib, DO NOT LOAD the
#   natbib package in this file, it is loaded automatically via the document
#   class glossa.cls)
# - [linguex] loads the linguex example package
#
# !!! A note on the use of linguex: in glossed examples, the third line of the
# example (the translation) needs to be prefixed with \glt. This is to allow a
# first line with the name of the language and the source of the example. See
# example (2) in the text for an illustration.
#
# !!! A note on the use of bibtex: for PhD dissertations to typeset correctly
# in the references list, the Address field needs to contain the city
# (for US cities in the format "Santa Cruz, CA")
#
bibliography: ["r-references.bib","refs.bib"]
biblio-style: "apalike"
link-citations: true
# The bibliography style is set automatically by glossa.cls when using
# either natbib or biblatex.
# Pandoc's citeproc is not supported.
#
abstract: |
   Behavioral measures of word-by-word reading time provide experimental evidence to test theories of language processing. A-maze is a recent method for measuring incremental sentence processing that can localize slowdowns related to syntactic ambiguities in individual sentences. We adapted A-maze for use on longer passages and tested it on the Natural Stories corpus. Participants were able to comprehend these longer text passages that they read via the Maze task.  Moreover, the Maze task yielded useable reaction time data with word predictability effects that were linearly related to surprisal, the same pattern found with other incremental methods. Crucially, Maze reaction times show a tight relationship with properties of the current word, with little spillover of effects from previous words. This superior localization is an advantage of Maze compared with other methods. Overall, we expanded the scope of experimental materials, and thus theoretical questions, that can be studied with the Maze task. 
# Specify keywords here:
keywords: stylesheet, glossa, article
# To add the word count, uncomment the following option and replace the number
# with the document word count.
# wordcount: 1000
output: 
  bookdown::pdf_book:
    base_format: rticles::glossa_article
    number_sections: yes
---

@@

gam1 <- read_rds(here("Analysis/models/bootstrapped_GAM_surprisal_predictions.rds")) %>%  ggplot( aes(x=surprisal, y=rt, ymin=CI_lower, ymax=CI_upper))+
  geom_line()+
  geom_ribbon(alpha=.3)+
  #facet_grid(s~model, scales="free_y")+
  facet_grid(s~model)+
  coord_cartesian(xlim=c(0,28))+
  labs(x="Surprisal (bits)", y="Effect on Reaction Time (ms)")+theme(axis.ticks.x=element_blank(), axis.title.x=element_blank(), axis.text.x=element_blank(), plot.margin=margin(t=0,r=0,b=0,l=0,unit="pt"))

all_mean_data <-  read_rds(here("Analysis/models/meaned_maze.rds"))

@@

```{r spr-gam, out.width="\\textwidth", fig.width=8, fig.height=4, fig.pos="ht", fig.cap="(ref:spr-gam-cap)"}
gam1 <- read_rds(here("Analysis/models/bootstrapped_spr_GAM_surprisal_predictions.rds")) %>%  ggplot( aes(x=surprisal, y=rt, ymin=CI_lower, ymax=CI_upper))+
  geom_line()+
  geom_ribbon(alpha=.3)+
  facet_grid(s~model)+
  coord_cartesian(xlim=c(0,28))+
  labs(x="Surprisal (bits)", y="Effect on Reaction Time (ms)")+theme(axis.ticks.x=element_blank(), axis.title.x=element_blank(), axis.text.x=element_blank(), plot.margin=margin(t=0,r=0,b=0,l=0,unit="pt"))

 all_mean_spr <-  read_rds(here("Analysis/models/meaned_spr.rds"))

@@

test_plot <- read_rds(here("Analysis/models/maze_spr_for_graph.rds"))

@@

```{r pre-error}
summ <- read_rds(here("Analysis/models/brms_maze_summ.rds")) %>% pivot_wider(names_from="model", values_from=c(`Estimate`)) %>% 
  mutate(Term=factor(Term, 
                        levels=c("Intercept", 
                                 "surprisal",
                                 "len",
                                 "freq",
                                 "surprisal:len",
                                 "len:freq",
                                 "prev_surp",
                                 "prev_len",
                                 "prev_freq",
                                 "prev_surp:prev_len",
                                 "prev_len:prev_freq")
                                 )) %>% 
  arrange(Term) %>% 
  mutate(Term=c("Intercept", "Surprisal", "Length", "Frequency",
                  "Surp x Length", "Freq x Length", "Past Surprisal",
                  "Past Length", "Past Freq", "Past Surp x Length", "Past Freq x Length"))

@@

summ_spr <- read_rds(here("Analysis/models/lmer_spr_summ.rds")) %>% 
  pivot_wider(names_from="model", values_from=c(`Estimate`)) %>% 
  mutate(Term=factor(Term, 
                        levels=c("(Intercept)", 
                                 "surp_center",
                                 "length_center",
                                 "freq_center",
                                 "surp_center:length_center",
                                 "length_center:freq_center",
                                 "past_surp_center",
                                 "past_length_center",
                                 "past_freq_center",
                                 "past_surp_center:past_length_center",
                                 "past_length_center:past_freq_center",
                                 "past2_surp_center",
                                 "past2_length_center",
                                 "past2_freq_center",
                                 "past2_surp_center:past2_length_center",
                                 "past2_length_center:past2_freq_center",
                                 "past3_surp_center",
                                 "past3_length_center",
                                 "past3_freq_center",
                                 "past3_surp_center:past3_length_center",
                                 "past3_length_center:past3_freq_center"
                                 )
                                 )) %>% 
  arrange(Term) %>% 
  mutate(Term=c("Intercept", "Surprisal", "Length", "Frequency","Surp x Length", "Freq x Length",
                "Past Surprisal", "Past Length", "Past Freq", "Past Surp x Length", "Past Freq x Length",
                "2Past Surprisal", "2Past Length", "2Past Freq", "2Past Surp x Length", "2Past Freq x Length",
                "3Past Surprisal", "3Past Length", "3Past Freq", "3Past Surp x Length", "3Past Freq x Length"))

@@

gam_linear <- read_rds(here("Analysis/models/gams_linearity_test.rds")) %>% 
  mutate(Term=factor(Term, 
                        levels=c("s(surprisal)",
                                 "ti(freq)",
                                 "ti(len)",
                                 "s(prev_surp)",
                                 "ti(prev_freq)",
                                 "ti(prev_len)",
                                 "(Intercept)",
                                 "surprisal",
                                 "prev_surp"
                                 )
                                 )) %>% 
  arrange(Term) %>% filter(Term %in% c("s(surprisal)", "s(prev_surp)", "surprisal","prev_surp")) %>% 
  mutate(Term=c("Spline Surprisal", "Spline Past Surprisal", "Linear Surprisal", "Linear Past Surprisal"))

@@

# Ethics and consent (optional) {#ethics-and-consent-optional .unnumbered}

Research involving human subjects, human material, or human data, must
have been performed in accordance with the Declaration of Helsinki.
Where applicable, the studies must have been approved by an appropriate
ethics committee. In the published paper, the authors should include a
statement detailing this approval, including the name of the ethics
committee and reference number of the approval. The identity of the
research subject(s) should be anonymised whenever possible. For research
involving human subjects, informed consent to participate in the study
must be obtained from participants (or their legal guardian).

# Funding information (optional) {#funding-information-optional .unnumbered}

Should the research have received a funding grant, then the grant
provider and grant number can be detailed in this section.

# Acknowledgements (optional) {#acknowledgements-optional .unnumbered}

@@

# Competing interests (mandatory) {#competing-interests-mandatory .unnumbered}

If any of the authors have any competing interests then these must be
declared. Consult the [Competing
Interests](http://www.glossa-journal.org/about/competinginterests/)
section on the Glossa website for more information. If there are no
competing interests to declare then the following statement should be
present: "The author(s) has/have no competing interests to declare".

# Authors' contributions (optional) {#contrib .unnumbered}

Here you can provide a sentence or a short paragraph detailing the
contribution of each author to the paper.

>>>>

---
title             : "A-maze of Natural Stories: Comprehension and surprisal in the Maze task"
shorttitle        : "A-maze of Natural Stories"
shortauthors: Veronica Boyce & Roger Levy
author: 
  - name          : "Veronica Boyce"
    affiliation   : "Stanford University"
    corresponding : yes    # Define only one corresponding author
    email         : "vboyce@stanford.edu"

  - name          : "Roger Levy"
    affiliation   : "Massachusetts Institute of Technology"

@@

classoption: [times, biblatex]
#
# Possible classoptions:
# - [times] for Times font (default if no option is chosen)
# - [cm] for Computer Modern font
# - [lucida] for Lucida font (not freely available)
# - [brill] open type font, freely downloadable for non-commercial use from
#   http://www.brill.com/about/brill-fonts; requires xetex
# - [charis] for CharisSIL font, freely downloadable from http://software.sil.org/charis/
# - for the Brill an CharisSIL fonts, you have to use the XeLatex
#   typesetting engine (not pdfLatex)
# - for headings, tables, captions, etc., Fira Sans is
#   used: https://www.fontsquirrel.com/fonts/fira-sans
# - [biblatex] for using biblatex (the default is natbib, DO NOT LOAD the
#   natbib package in this file, it is loaded automatically via the document
#   class glossa.cls)
# - [linguex] loads the linguex example package
#
# !!! A note on the use of linguex: in glossed examples, the third line of the
# example (the translation) needs to be prefixed with \glt. This is to allow a
# first line with the name of the language and the source of the example. See
# example (2) in the text for an illustration.
#
# !!! A note on the use of bibtex: for PhD dissertations to typeset correctly
# in the references list, the Address field needs to contain the city
# (for US cities in the format "Santa Cruz, CA")
#
bibliography: ["r-references.bib","refs.bib"]
biblio-style: "apalike"
link-citations: true
# The bibliography style is set automatically by glossa.cls when using
# either natbib or biblatex.
# Pandoc's citeproc is not supported.
#
abstract: |
   Behavioral measures of word-by-word reading time provide experimental evidence to test theories of language processing. A-maze is a recent method for measuring incremental sentence processing that can localize slowdowns related to syntactic ambiguities in individual sentences. We adapted A-maze for use on longer passages and tested it on the Natural Stories corpus. Participants were able to comprehend these longer text passages that they read via the Maze task.  Moreover, the Maze task yielded useable reaction time data with word predictability effects that were linearly related to surprisal, the same pattern found with other incremental methods. Crucially, Maze reaction times show a tight relationship with properties of the current word, with little spillover of effects from previous words. This superior localization is an advantage of Maze compared with other methods. Overall, we expanded the scope of experimental materials, and thus theoretical questions, that can be studied with the Maze task. 
# Specify keywords here:
keywords: A-Maze, reading time measure, surprisal,
# To add the word count, uncomment the following option and replace the number
# with the document word count.
# wordcount: 1000
output: 
  bookdown::pdf_book:
    base_format: rticles::glossa_article
    number_sections: yes
---

@@

gam1 <- read_rds(here("Analysis/models/bootstrapped_GAM_surprisal_predictions.rds")) %>% mutate(model=factor(model, 
                        levels=c("5-gram",
                                 "GRNN",
                                 "Transformer-XL",
                                 "GPT-2"))) %>%
 ggplot( aes(x=surprisal, y=rt, ymin=CI_lower, ymax=CI_upper))+
  geom_line()+
  geom_ribbon(alpha=.3)+
  #facet_grid(s~model, scales="free_y")+
  facet_grid(s~model)+
  coord_cartesian(xlim=c(0,28))+
  labs(x="Surprisal (bits)", y="Effect on Reaction Time (ms)")+theme(axis.ticks.x=element_blank(), axis.title.x=element_blank(), axis.text.x=element_blank(), plot.margin=margin(t=0,r=0,b=0,l=0,unit="pt"))

all_mean_data <-  read_rds(here("Analysis/models/meaned_maze.rds")) %>% mutate(model=factor(model, 
                        levels=c("5-gram",
                                 "GRNN",
                                 "TXL",
                                 "GPT-2")))

@@

```{r spr-gam, out.width="\\textwidth", fig.width=8, fig.height=4, fig.pos="ht", fig.cap="(ref:spr-gam-cap)"}
gam1 <- read_rds(here("Analysis/models/bootstrapped_spr_GAM_surprisal_predictions.rds")) %>%
  mutate(model=factor(model, 
                        levels=c("5-gram",
                                 "GRNN",
                                 "Transformer-XL",
                                 "GPT-2"))) %>%
  ggplot( aes(x=surprisal, y=rt, ymin=CI_lower, ymax=CI_upper))+
  geom_line()+
  geom_ribbon(alpha=.3)+
  facet_grid(s~model)+
  coord_cartesian(xlim=c(0,28))+
  labs(x="Surprisal (bits)", y="Effect on Reaction Time (ms)")+theme(axis.ticks.x=element_blank(), axis.title.x=element_blank(), axis.text.x=element_blank(), plot.margin=margin(t=0,r=0,b=0,l=0,unit="pt"))

all_mean_spr <-  read_rds(here("Analysis/models/meaned_spr.rds")) %>% mutate(model=factor(model, 
                        levels=c("5-gram",
                                 "GRNN",
                                 "TXL",
                                 "GPT-2")))

@@

test_plot <- read_rds(here("Analysis/models/maze_spr_for_graph.rds"))%>% mutate(model=factor(model, 
                        levels=c("5-gram",
                                 "GRNN",
                                 "TXL",
                                 "GPT-2")))

@@

```{r pre-error}
summ <- read_rds(here("Analysis/models/brms_maze_summ.rds")) %>%
  pivot_wider(names_from="model", values_from=c(`Estimate`)) %>% 
  mutate(Term=factor(Term, 
                        levels=c("Intercept", 
                                 "surprisal",
                                 "len",
                                 "freq",
                                 "surprisal:len",
                                 "len:freq",
                                 "prev_surp",
                                 "prev_len",
                                 "prev_freq",
                                 "prev_surp:prev_len",
                                 "prev_len:prev_freq")
                                 )) %>% 
  arrange(Term) %>% 
  mutate(Term=c("Intercept", "Surprisal", "Length", "Frequency",
                  "Surp x Length", "Freq x Length", "Past Surprisal",
                  "Past Length", "Past Freq", "Past Surp x Length", 
                "Past Freq x Length")) %>%
  select(Term, `5-gram`, GRNN, `Transformer-XL`=TXL,`GPT-2`)

@@

summ_spr <- read_rds(here("Analysis/models/lmer_spr_summ.rds")) %>% 
  pivot_wider(names_from="model", values_from=c(`Estimate`)) %>% 
  mutate(Term=factor(Term, 
                        levels=c("(Intercept)", 
                                 "surp_center",
                                 "length_center",
                                 "freq_center",
                                 "surp_center:length_center",
                                 "length_center:freq_center",
                                 "past_surp_center",
                                 "past_length_center",
                                 "past_freq_center",
                                 "past_surp_center:past_length_center",
                                 "past_length_center:past_freq_center",
                                 "past2_surp_center",
                                 "past2_length_center",
                                 "past2_freq_center",
                                 "past2_surp_center:past2_length_center",
                                 "past2_length_center:past2_freq_center",
                                 "past3_surp_center",
                                 "past3_length_center",
                                 "past3_freq_center",
                                 "past3_surp_center:past3_length_center",
                                 "past3_length_center:past3_freq_center"
                                 )
                                 )) %>% 
  arrange(Term) %>% 
  mutate(Term=c("Intercept", "Surprisal", "Length", "Frequency","Surp x Length", "Freq x Length",
                "Past Surprisal", "Past Length", "Past Freq", "Past Surp x Length", "Past Freq x Length",
                "2Past Surprisal", "2Past Length", "2Past Freq", "2Past Surp x Length", "2Past Freq x Length",
                "3Past Surprisal", "3Past Length", "3Past Freq", "3Past Surp x Length", "3Past Freq x Length")) %>%
  select(Term, `5-gram`, GRNN, `Transformer-XL`=TXL,`GPT-2`)

@@

gam_linear <- read_rds(here("Analysis/models/gams_linearity_test.rds")) %>% 
  mutate(Term=factor(Term, 
                        levels=c("s(surprisal)",
                                 "ti(freq)",
                                 "ti(len)",
                                 "s(prev_surp)",
                                 "ti(prev_freq)",
                                 "ti(prev_len)",
                                 "(Intercept)",
                                 "surprisal",
                                 "prev_surp"
                                 )
                                 )) %>% 
  arrange(Term) %>% filter(Term %in% c("s(surprisal)", "s(prev_surp)", "surprisal","prev_surp")) %>% 
  mutate(Term=c("Spline Surprisal", "Spline Past Surprisal", "Linear Surprisal", "Linear Past Surprisal")) %>%
  select(Term, `5-gram`, GRNN, `Transformer-XL`=TXL,`GPT-2`)

@@

<!--# Ethics and consent (optional) {#ethics-and-consent-optional .unnumbered}

Research involving human subjects, human material, or human data, must
have been performed in accordance with the Declaration of Helsinki.
Where applicable, the studies must have been approved by an appropriate
ethics committee. In the published paper, the authors should include a
statement detailing this approval, including the name of the ethics
committee and reference number of the approval. The identity of the
research subject(s) should be anonymised whenever possible. For research
involving human subjects, informed consent to participate in the study
must be obtained from participants (or their legal guardian).
-->
<!--# Funding information (optional) {#funding-information-optional .unnumbered}

Should the research have received a funding grant, then the grant
provider and grant number can be detailed in this section.-->

# Acknowledgements {#acknowledgements-optional .unnumbered}

@@

# Competing interests {#competing-interests-mandatory .unnumbered}

The authors have no competing interests to declare.

# Authors' contributions {#contrib .unnumbered}
Veronica Boyce contributed Conceptualization, Formal Analysis, Investigation, Methodology, Software, and Writing - Original Draft Preparation. Roger Levy contributed Conceptualization, Formal Analysis, Funding Acquisition, Methodology, Supervision, and Writing - Review & Editing.
